Effects of bariatric surger and weight loss on resting-state functional connectivity

### 2020-12-03 manuscript/intro_rsfmri.Rmd
corrected mistakes

<!-- Reward network -->
Little is known about the potential beneficial effects of bariatric surgery on brain function in the obese population. Thus, one recently emerging research focus is how brain metabolism and functional connectivity changes in response to surgery, in particular with regard to hedonic motivations for food, which play a fundamental role in weight relapse.
The reward network comprising the ventromedial prefrontal cortex (vmPFC), the striatum (STR) but also the anterior insula (antINS) has been linked to food valuation processes in decision-making in lean individuals [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018; @Wiemerslage_2016]. These regions also seem to be altered after bariatric surgery [@Lips_2014; @Hogenkamp_2016; @Garcia_2012]. <!-- Zitationen checken -->
<!-- Normalization of connectivity --> Obese individuals exhibited increased connectivity between frontal and mesolimbic regions and weaker connectivity within regions of the mesolimbic system (encompassing limbic and striatal regions). But these differences seemed to normalize, as trends reversed when looking at the resting state connectivity after compared to before the surgery [@Duan_2020]. Resting state activitiy increased in subcortical regions, such as the insula and striatum<!-- specifically putamen --> [@Wiemerslage_2016]. 
<!-- mechanism reward and appetite, overeating --> Interestingly, patients report changes in food preferences and eating behavior after the surgery that may influence weight loss beyond malabsorption of nutrients and restricted food ingestion. Those psychophysiological and behavioral effects are presumably related to post-surgical changes in visceral sensory perceptions and endocrinological signaling. Corresponding information is processed by the fronto-mesolimbic system and affects appetite and reward signalling which may influence weight regulation [@Mulla_2017; @Nudel_2019].
<!-- results in lean -->

@@

<!-- Default mode network --> 
Because multiple resting state networks are relevant to obesity and associated with abnormal neurocircuitry interactions [@Ding_2020], extensive research has been devoted to investigating the the relationship between the control network and the reward network. However, the dynamics within the default network may also be of importance because it is associated with processing of self-referent as well as introspective information, such as hunger sensation [@;@] and appetite [@Tregellas_2011]. <!-- parietal activation correlated with appetite measures after an overnight fast [@Tregellas_2011]. --> 
Weight regulation and energy expenditure, on the other hand, are highly dependent on a complex interplay of external and internal sensory processing, as well as molecular homeostasis regulation. 
<!-- Obesity = decompensation of the psychophysiological system that ensures homeostasis. -->

The default mdoe network is mainly composed of the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and inferior and lateral parietal cortex (approx. BA 39) [@Raichle_2015]. The right ventral anterior cingulate cortex (vACC), superior frontal cortex, and inferior temporal cortex show weaker but also reliable association with the PCC over the course of one year. Even broader conceptions count left VACC, middle temporal gyrus, and parahippocampal cortex, although associations with the PCC are not as stable [@Chou_2012].
<!-- study results --> 
Functional connectivity of this network has been shown to be linked to BMI [@Beyer_2017] and altered after bariatric surgery [@Olivo_2017].
Generally, studies show an increased activation in the default mode network, especially in the precuneus and posterior cingulate cortex (PCC) in obese individuals [@Kullmann_2011; @].
Comparing spontaneous activitiy before and after surgery shows a decrease in areas including superior frontal regions<!-- gehört zu SFC -->, and a decrease superior parietal lobule when in sated state [@Wiemerslage_2016].
Concerning spontaneuous functional connectivity in these regions, patient having undergone bariatric surgery resemble normal-weight individuals rather than obese individual in having a weaker connectivity within the default mode network [@Frank_2013].
A longitudinal study investigating confirmed that while there was stronger connectivity in the default model network <!-- (PCC with anterior STG and anterior temporal fusiform gyrus and the Precuneus with the anterior STG) --> in obese patients before the surgery, it seemed weakened over the course of one year. Because differences between lean controls and obese individuals that existed prebariatric surgery were no longer detectable after the surgery, the authors argue for a normalization in functional connectivity<!-- and PCC/PC connectivity with anterior STG are most affected by the surgery -->[@Olivo_2017].
Consistent with these findings, other studies reported a reduction of functional connectivity of these and other regions of the default mode network, such as the vmPFC <!-- with the dorsolateral prefrontal cortex (DLPFC) --> after the surgery [@Li_2018].

@@

<!--  -->

<!-- novelty statement --> As described previously, several studies have investigated the effects of bariatric surgery on the intrinsic functional connectivity in patients. Yet, to our knowledge there is no quasi-experimental study, comparing a cohort of patients to a cohort with similar characteristics but unexposed to the intervention, i.e. bariatric surgery, over an extended time period as would be demanded to draw more substantiated inference [@George_2016].
<!-- hypotheses --> In this study, we aim to bridge this gap and investigate possible dynamics of within-reward network connectivity before and after bariatric surgery by comparison with a contemporaneously measured waiting-control group. Moreover, we will explore whether the changes in functional connectivity are associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential gradient link. In addition, we will investigate potential changes in within-network functional connectivity after surgery in other resting-state networks such as default mode network (DMN).
Participants are measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008]. Notably, we will consider that different courses in increase and decrease of functional connectivity that are not necessarily linear are possible over the course of one year, depending on the phase of weight management [@Olivo_2017].

>>>>

<!-- Reward network -->
Little is known about the potential beneficial effects of bariatric surgery on brain function in the obese population. Thus, one recently emerging research focus is how brain metabolism and functional connectivity changes in response to surgery, in particular with regard to hedonic motivations for food, which play a fundamental role in weight relapse.
The reward network comprising the ventromedial prefrontal cortex (vmPFC), the striatum (STR) but also the anterior insula (antINS) has been linked to food valuation processes in decision-making in lean individuals [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018; @Wiemerslage_2016]. These regions also seem to be altered after bariatric surgery [@Lips_2014; @Hogenkamp_2016; @Garcia_2012]. <!-- mehr zu Zitationen -->
<!-- Normalization of connectivity --> Obese individuals exhibited increased connectivity between frontal and mesolimbic regions and weaker connectivity within regions of the mesolimbic system (encompassing limbic and striatal regions). But these differences seemed to normalize, as trends reversed when looking at the resting state connectivity after compared to before the surgery [@Duan_2020]. Resting state activitiy increased in subcortical regions, such as the insula and striatum<!-- specifically putamen --> [@Wiemerslage_2016]. 
<!-- mechanism reward and appetite, overeating --> Interestingly, patients report changes in food preferences and eating behavior after the surgery that may influence weight loss beyond malabsorption of nutrients and restricted food ingestion. Those psychophysiological and behavioral effects are presumably related to post-surgical changes in visceral sensory perceptions and endocrinological signaling. Corresponding information is processed by the fronto-mesolimbic system and affects appetite and reward signalling which may influence weight regulation [@Mulla_2017; @Nudel_2019].
<!-- results in lean -->

@@

<!-- Default mode network --> 
Because multiple resting state networks are relevant to obesity and associated with abnormal neurocircuitry interactions [@Ding_2020], extensive research has been devoted to investigating the relationship between the control network and the reward network. However, the dynamics within the default network may also be of importance because it is associated with processing of self-referent as well as introspective information, such as hunger sensation [@;@] and appetite [@Tregellas_2011]. <!-- parietal activation correlated with appetite measures after an overnight fast [@Tregellas_2011]. --> 
Weight regulation and energy expenditure, on the other hand, are highly dependent on a complex interplay of external and internal sensory processing, as well as molecular homeostasis regulation. 
<!-- Obesity = decompensation of the psychophysiological system that ensures homeostasis. -->

The default mode network is mainly composed of the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and inferior and lateral parietal cortex (approx. BA 39) [@Raichle_2015]. The right ventral anterior cingulate cortex (vACC), superior frontal cortex, and inferior temporal cortex show weaker but also reliable association with the PCC over the course of one year. Even broader conceptions count left vACC, middle temporal gyrus, and parahippocampal cortex, although associations with the PCC are not as stable [@Chou_2012].
<!-- study results --> 
Functional connectivity of this network has been shown to be linked to BMI [@Beyer_2017] and altered after bariatric surgery [@Olivo_2017].
Generally, studies show an increased activation in the default mode network, especially in the precuneus and posterior cingulate cortex (PCC) in obese individuals [@Kullmann_2011; @].
Comparing spontaneous activitiy before and after surgery shows a decrease in areas including superior frontal regions<!-- gehört zu SFC -->, and a decrease superior parietal lobule when in sated state [@Wiemerslage_2016].
Concerning spontaneuous functional connectivity in these regions, patient having undergone bariatric surgery resemble normal-weight individuals rather than obese individual in having a weaker connectivity within the default mode network [@Frank_2013].
A longitudinal study investigating confirmed that while there was stronger connectivity in the default model network <!-- (PCC with anterior STG and anterior temporal fusiform gyrus and the Precuneus with the anterior STG) --> in obese patients before the surgery, it seemed weakened over the course of one year. Because differences between lean controls and obese individuals that existed prebariatric surgery were no longer detectable after the surgery, the authors argue for a normalization in functional connectivity<!-- and PCC/PC connectivity with anterior STG are most affected by the surgery -->[@Olivo_2017].
Consistent with these findings, other studies reported a reduction of functional connectivity of these and other regions of the default mode network, such as the vmPFC <!-- with the dorsolateral prefrontal cortex (DLPFC) --> after the surgery [@Li_2018].

@@

<!-- novelty statement --> As described previously, several studies have investigated the effects of bariatric surgery on the intrinsic functional connectivity in patients. Yet, to our knowledge there is no quasi-experimental study, comparing a cohort of patients to a cohort with similar characteristics but unexposed to the intervention, i.e. bariatric surgery, over an extended time period as would be demanded to draw more substantiated inference [@George_2016].
<!-- hypotheses --> In this study, we aim to bridge this gap and investigate possible dynamics of within-reward network connectivity before and after bariatric surgery by comparison with a contemporaneously measured waiting-control group. Moreover, we will explore whether the changes in functional connectivity are associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential gradient link. In addition, we will investigate potential changes in within-network functional connectivity after surgery in other resting-state networks such as default mode network.
Participants are measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008]. Notably, we will consider that different courses in increase and decrease of functional connectivity that are not necessarily linear are possible over the course of one year, depending on the phase of weight management [@Olivo_2017].
